---
title: Analyses of 27 years of stand structure development and basal area growth response to a range of initial basal area density levels, 1992-2019 
author: "Mike Jull and Hardy Griesbauer"
date: "29/03/2020"
output:
  word_document:
    reference_docx: wordTemplate2.docx
  html_document:
    df_print: paged
csl: cjfr.csl
bibliography: ep1162Bib.bib
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)

# Load libraries
library(tidyverse)
library(here)
library(lme4)
library(lmerTest)
library(broom)
library(MuMIn)
library(ggeffects)
library(car)
library(flextable)
library(captioner)
library(lsmeans)
library(cowplot)

# Load data
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# set up captions

# Code to create captions for tables
tableNums <- captioner(prefix = "Table")

# Code to create captions for figures
figNums <- captioner(prefix="Figure")

# Code to create captions for appendices
appNums <- captioner(prefix="Appendix")

# Set ggplot theme
theme_set(theme_bw())

# Some additional data prep

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)
dat<- # assign variable
  dat %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% 
  mutate(Plot=as.factor(Plot))

# formatting plotDat
plotDat<-
  plotDat %>% 
  mutate(Plot=as.factor(Plot))

```



# Methods

## Analysis

### Tree-level growth responses
To test the hypothesis that tree-level growth response over the 1992-2019 period varied with interactions between (i) tree basal area at the time of treatment (1992); (ii) residual basal area; and (iii) species, we fit linear mixed effects models with log-transformed 1992-2019 basal area increment as the response variable, and treatment, species and log-transformed 1992 basal area as interacting fixed effects, and plot as a random intercept effect.  We also fit a similar model with the 1992-2019 basal area increment expressed as a percentage of 1992 basal area.  Plots of residuals and fit lines with orginal data were examined to evaluate model goodness of fit.  Variance explained for each model are reported using the approach described in Nakagawa et al. [-@Nakagawa2013], as implemented in the MuMIn package for R [@add later].

We also examined how basal area increment at the tree-level varied over the four measurement periods, by fitting a similar model as described above, with an additional fixed effect of measurement period.  This allowed us to examine temporal changes in growth responses, including evaluating for a potential lag effect.  We do note that because we did not measure increment before treatment, we cannot address questions regarding short-term growth declines associated with treatment.


## Stand structure and growth dynamics

### Changes in tree density over time
We qualitatively assessed changes in tree density over time through density diagrams that displayed tree density (stems per hectare) by diameter class, faceted by measurement period, species and treatment unit.  The effects of tree density changes on basal area was also qualitatively evaluated by summarizing basal area per hectare by diameter class, species and treatment unit.

### Stand-level basal area and volume growth


#### Volume estimates
Tree volumes were estimated using species- and region-specific equations [@nigh2016] based on tree height and DBH.  We estimated tree heights by fitting a single diameter-height models developed with a linear mixed effects model, with log-transformed DBH as a fixed effect, and tree nested within plot as a random effect.  Trees with broken tops were omitted from diameter-height models. 



# Results



### Tree-level growth responses
```{r tree-level growth responses}

# Let's filter only for trees that have measurements in 1992 and 2019 
treeList<- # assign variable
  dat %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Sx","Bl")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  filter(Year %in% c(1992,2019)) %>% #filter for start and end dates
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==2) %>% #filter for trees with 5 measurements
  pull(TreeID)
  

# derive periodic increment (period diameter or basal area increment)
piDat<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,BA) %>% 
  pivot_wider(names_from=Year,values_from=BA) %>% 
  mutate(PI=`2019`-`1992`) %>% # 
  filter(PI>0) %>% # one tree has a PI of 0, has to be removed
  mutate(PIpct=PI/`1992`*100) %>% # express periodic increment a a percentage
  dplyr::select(Plot,TreeID,Species,PI,PIpct,Init=`1992`) %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(BA.Target=paste("TU:",BA.Target,sep="")) # create facet column

# Create a table for the appendix with growth rates
piDat.table<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>%
  filter(Year %in% c(1992,2019)) %>% 
  # mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  dplyr::select(Plot,TreeID,Species,Year,DBH) %>% 
  pivot_wider(names_from=Year,values_from=DBH) %>% 
  mutate(BA92=(`1992`/2/100)^2*pi) %>% 
  mutate(BA19=(`2019`/2/100)^2*pi) %>%
  mutate(PI=BA19-BA92) %>% # 
  filter(PI>0) %>%   # one tree has a PI of 0, has to be removed
  mutate(initDiam=`1992`) %>% 
  mutate(diamClass=cut(initDiam,breaks=c(7.5,12.5,18.5,30,90),include.lowest=T,right=F)) %>% 
  left_join(plotDat,by="Plot") %>% 
  mutate(BA.Target=paste("TU:",BA.Target,sep="")) %>%  # create facet column
  group_by(BA.Target,Species,diamClass) %>% 
  summarise(BAImean=mean(PI),
            BAIse=sd(PI)/sqrt(length(PI))) %>% 
  arrange(BA.Target,diamClass,Species)
  
  
# Create figures of raw data to put into Appendix
pi.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PI),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target)  +
   ylab("log(BA increase, m²)")+
  xlab("log(Tree BA in 1992, m²)")+
  theme(legend.position = "none")


piPCT.Init.rawPlot<-  
  piDat %>% 
  ggplot(aes(x=log(Init),y=log(PIpct),color=Species))+
  geom_point()+
  geom_smooth(method="lm",se=FALSE,color="black")+
  facet_grid(Species~BA.Target) +
  ylab("log(% BA increase)")+
  xlab("log(Tree BA in 1992, m²)")+
  theme(legend.position="none")


# fixed effects model with species pooled, using PBAI
fitBA<-
  piDat %>% 
  lmer(log(PI)~BA.Target:log(Init):Species+(1|Plot),data=.)

# fixed effects model with species pooled, using PBAIpct
fitBApct<-
  piDat %>% 
  lmer(log(PIpct)~BA.Target:log(Init):Species+(1|Plot),data=.)

# Create figure showing marginal means from model
fitBA.plot<-
  ggpredict(fitBA,terms=c("Init [exp]","Species","BA.Target")) %>% 
  dplyr::rename(Species=group) %>% 
  ggplot(aes(x,predicted,color=Species))+
  geom_line()+
  facet_wrap(~facet)+
  ylab("Predicted 27-year basal area increment (m²)")+
  xlab("Tree BA in 1992 (m²)")+
  theme_bw()+
  theme(legend.position="bottom")
  

# Create figure showing marginal means from model
fitBApct.plot<-
  ggpredict(fitBApct,terms=c("Init [exp]","Species","BA.Target")) %>% 
  dplyr::rename(Species=group) %>% 
  ggplot(aes(x,predicted,color=Species))+
  geom_line()+
  facet_wrap(~facet)+
  ylab("Predicted 27-year basal area % increase")+
  xlab("Tree BA in 1992 (m²)")+
  theme(legend.position="bottom")

  


# Captions
appNums(name="piInitRaw",display=FALSE)
appNums(name="piPCTInitRaw",display=FALSE)
appNums(name="piDataRaw",display=FALSE)

figNums(name="fitBA.plot",display=FALSE)
figNums(name="fitBApct.plot",display=FALSE)

```

A comparison of tree-level 1992-2019 basal area increment to tree size at the time of treatment showed that basal area increment over 27 years increased with tree size, and that this relationship varied between species and among treatment units (`r appNums("piInitRaw",display="cite")`).  A linear mixed effects model using 
`r nrow(piDat)`
 trees (`r nrow(filter(piDat,Species=="Sx"))`
spruce, 
`r nrow(filter(piDat,Species=="Bl"))` fir) confirmed that slopes and intercepts varied with species and residual basal area (p<0.001 for all interactions).  This model explained 
`r round(r.squaredGLMM(fitBA)[1]*100,1)`
% variation (marginal R^2^) in diameter increment over the 27 years of the study, and showed that diameter increment increased with initial tree size, and increment was higher in the treatment units with basal area removed, compared to the control, and that spruce trees had higher basal area increment than fir (`r figNums("fitBA.plot",display="cite")`).  A similar model predicting basal area increment as a percentage of tree size at the time of treatment showed that smaller trees grew proportionally more than larger trees, with a strong negative exponential decrease with initial tree size (`r appNums("piPCTInitRaw",display="cite")` and `r figNums("fitBApct.plot",display="cite")`), however, the model fit was relatively poorer, explaining 
`r round(r.squaredGLMM(fitBApct)[1]*100,1)`
% of variation in % basal area increase.  

```{r tree level time lag response}
treeList<- # assign variable
  dat %>% 
  filter(TreeID!="New") %>% #remove new trees
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(Species %in% c("Sx","Bl")) %>% # filter birch and fdi out of dataset
  filter(Status=="Live") %>% # filter for live trees
  group_by(TreeID) %>% 
  summarise(N=n()) %>% 
  filter(N==5) %>% #filter for trees with 5 measurements
  pull(TreeID)

# derive mean annual increment by period

piDat<-
  dat %>% 
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create unique treeID
  filter(TreeID %in% treeList) %>% 
   left_join(plotDat,by="Plot") %>% 
  mutate(BA=(DBH/2/100)^2*pi) %>% # convert DBH to basal area
  dplyr::select(BA.Target,Plot,TreeID,Species,Year,val=BA) %>% 
  pivot_wider(names_from=Year,values_from=val) %>% 
  mutate(BA.Target=paste("TU:",BA.Target,sep="")) %>%  # create facet column
  
  # Create growth columns
  mutate(`2019`=(`2019`-`2009`)/11) %>% # divide by 11 to get annual increment
  mutate(`2009`=(`2009`-`1997`)/11) %>% # divide by 11 to get annual increment
  mutate(`1997`=(`1997`-`1994`)/3) %>% # divide by 3 to get annual increment
  mutate(`1994`=(`1994`-`1992`)/3) %>% # divide by 3 to get annual increment
  mutate(`Init`=`1992`) %>% # create initial size column
  pivot_longer(cols=c(`1992`,`1994`,`1997`,`2009`,`2019`),values_to="Inc",names_to="Year") %>% 
  filter(Year!=1992) %>% 
  filter(Inc>0) %>% # need to remove any trees with increment of 0- %>% 
  mutate(initClass=cut(Init,breaks=seq(from=7.5,to=90,by=20))) %>% 
  mutate(incPCT=Inc/Init*100) 
 

# Raw plots
baAnn.raw.plot<-
  # Plot boxplot of increment by year
  piDat %>% 
  ggplot(aes(x=Year,y=Inc,color=BA.Target))+
  geom_boxplot()+
  facet_grid(Species~BA.Target)+
  xlab("Measurement year")+
  ylab("Annualized basal area increment (m²)")+
  theme(legend.position = "none")+
  theme(legend.position = "none")


# Fit lmer
# fixed effects model
fitBA.Ann<-
  piDat %>% 
  lmer(log(Inc)~Year+Species+BA.Target+ (1|Plot/TreeID) ,data=.)
 
 # Create figure showing marginal means from model
fitBA.Ann.plot<-
  ggpredict(fitBA.Ann,terms=c("Year","Species","BA.Target"))  %>%
  plot()+
  theme_bw()+
  ylab("Predicated annualized basal area increment (m²)")+
  xlab("Measurement year")+
  ggtitle("")+
  theme(legend.position = "bottom")
  
# Captions 
appNums(name="baAnnRaw",display=FALSE)
figNums(name="fitBA.Ann.plot",display=FALSE)

```

To detect differences in growth rates over the 27-year period of this study, we compared growth rates between the four measurement periods, by species and treatment unit (`r appNums(name="baAnnRaw",display="cite")`).  Growth rates were lowest in the three years immediately after treatment, and increased over the subsequent two measurement periods (1994-1997 and 1997-2009), before decreasing again in the last measurement period (2009-2019).  This temporal pattern appeared consistent between species and among treatment units; as well, spruce trees had higher growth rates than fir over all periods and treatment units, and growth rates were higher in the high-removal treatment unit, compared to the low-removal treatment and control units.  A linear mixed effects model with species, measurement year and treatment unit as fixed effects (not shown) explained `r MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100`% variation (marginal R^2^) in log-transformed basal area increment (m^2^ at breast height) over different periods (`r figNums("fitBA.Ann.plot",display="cite")`).

### Tree density
```{r sph by diam class species and time}

# need to create  a list of plots with five measurements
plotDenom<-
  dat %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  group_by(BA.Target,Year,Plot) %>% 
  summarise(Num.Plots=n()) %>% 
  ungroup() %>% 
  dplyr::select(BA.Target:Plot) %>% 
  group_by(BA.Target,Year) %>% 
  summarise(Num.Plots=n())
 

# PLOTS
# Raw plot - examine SPH in each plot to detect where mortality occurred
# sphPlot.raw<-
#   dat %>% 
#   left_join(plotDat,by="Plot") %>% # join with plot data
#   filter(Status=="Live") %>% # filter for live trees only
#   group_by(BA.Target,Species,Plot,Year,DBH.class) %>% 
#   summarise(SPH=n()*20)  %>% 
#   filter(Species %in% c("Bl","Sx")) %>% # remove Fdi and Ep
#   ggplot()+
#   aes(x=DBH.class,y=SPH,group=Species,color=Species)+
#   geom_point()+
#   facet_grid(Plot~Year)+
#   theme(axis.text.x = element_text(angle = 90, hjust = 1),
#         legend.position="bottom")+
#   xlab("DBH Class (cm)")
# 
# appNums(name="sphRawPlot",display=FALSE)

# 1. SPH by diameter class and species for 1992 and 2019 periods
sphPlot1<-
  dat %>% 
  filter(Year %in% c(1992,2019)) %>% # Select for two time periods
    filter(Species %in% c("Bl","Sx")) %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  group_by(BA.Target,Year,Species,DBH.class) %>% 
  summarise(SPH.Sum=n()*20) %>% # plot multiplier
  left_join(plotDenom,by=c("BA.Target","Year")) %>% 
  mutate(SPH=SPH.Sum/Num.Plots) %>% 
  
  ggplot()+
  aes(x=DBH.class,y=SPH,group=Species,fill=Species,color=Species)+
  geom_point()+
  geom_path()+
    # geom_density(alpha=0.4)+
  facet_grid(Year~BA.Target)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position = "bottom")+
  xlab("DBH Class (cm)")
  
  figNums(name="sphSpeciesPlot",display=FALSE)
```
  
```{r plot pct change in SPH by diam class between 92 and 19}  

# % change in diameter class between 1992 and 2019 time periods by species
 sphPCTPlot<- 
  dat %>% 
  filter(Year %in% c(1992,2019)) %>% 
  filter(Status=="Live") %>% 
  filter(Species %in% c("Bl","Sx")) %>% 
  group_by(Plot,Species,Year,DBH.class) %>% 
  summarise(SPH.Total=n()) %>% 
  pivot_wider(names_from="Year",values_from="SPH.Total") %>% 
  mutate(Pct.Change=(`2019`-`1992`)/`1992`*100) %>% 
  mutate(Pct.Change=round(Pct.Change,3)) %>%
  drop_na(Pct.Change) %>% 
  left_join(plotDat,by="Plot") %>% # join with plot data
  
  # Plotting
  ggplot()+
  aes(x=DBH.class,y=Pct.Change,fill=Species)+
  geom_boxplot()+
  geom_hline(yintercept=0, linetype="dashed", color="darkgrey")+
  facet_grid(Species~BA.Target)+
  theme(axis.text.x = element_text(angle = 90, hjust = 1,vjust=0.5),
        legend.position="bottom")+
  xlab("DBH Class (cm)")+
  ylab("% change in SPH between 1992 and 2019")
  

  figNums(name="sphPCTPlot",display=FALSE)


# ggsave(filename="baDiamClass.png")


```

Mean plot tree density by diameter class diagrams showed that across all treatment levels, subalpine fir comprised the majority of stems in the smaller diameter classes, both in the year after treatment, and in 2019 (`r figNums("sphSpeciesPlot",display="cite")`). The diameter class distribution of fir in all treatment units was negative exponential.  Mean spruce density was lower than fir, with a relatively flatter diameter distribution.  Within the high-removal treatment unit, fir density increased mostly in the smaller diameter classes between 1992 and 2019; in the low-removal treatment unit, fir density increased most in trees between 12.5 to 27.5cm DBH; and in the control treatment unit, fir density increased mostly in trees above 22.5cm DBH ( `r figNums("sphPCTPlot",display="cite")`).  Spruce tree density increased less than fir, with the exception of an increase of over 250% in SPH in the smallest diameter class in the light-removal treatment unit.

## Stand-level changes

### Volume estimates


### Height-diameter models

```{r height diameter raw plot}

heightDiamRaw.plot<-
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
    
    # Plotting
  ggplot()+
  aes(x=DBH,y=Height,fill=Species)+
  geom_point()+
  facet_grid(Year~Species)+
  theme(legend.position = "none")+
  ylab("Height (m)")+
  xlab("DBH (cm)")+
  geom_smooth(formula = y~log(x),method="lm",se=FALSE)

# Fit model
htDiamFit<- 
  dat %>% 
    filter(Species %in% c("Bl","Sx")) %>% # filter for Bl and Sx for now
    filter(Status=="Live") %>% # filter for live trees
    filter(is.na(DBT)) %>% # filter for trees with no broken tops
    drop_na(Height) %>% # remove trees where height was not measured
  mutate(TreeID=paste(Plot,TreeID,sep="-")) %>% # create tree ID column
  lmer(log(Height)~log(DBH)+(1|Plot/TreeID),data=.)

# Apply fit model to data and create new height column
dat<-
  dat %>% 
  mutate(Height.Pred=coalesce(Height,exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))))

# Height model fit plot
htModelFit.plot<-
  dat %>% 
  mutate(h2=exp(fixef(htDiamFit)[1]+fixef(htDiamFit)[2]*log(DBH))) %>% 
  ggplot(aes(x=Height,y=h2))+
  geom_point()+
  xlim(0,40)+
  ylim(0,40)+
  geom_smooth(method="lm",se=F)+
  ylab("Predicted height (m) from diameter-height model")+
  xlab("Measured height (m)")

# create captions
appNums("heightDiamRawPlot",display="FALSE")
appNums("htModelFit",display="FALSE")

```

Tree heights had a positive nonlinear relationship with diameter, for both species and across all time periods (`r appNums("heightDiamRawPlot",display="cite")`).  A comparison of height-diameter models showed that intercepts and slopes of log-transformed heights did not vary significantly (p<0.05) between species or time periods (not shown), therefore a single height-diameter linear mixed-effects model was developed using `r dat %>% filter(Species %in% c("Bl","Sx")) %>% filter(Status=="Live") %>% filter(is.na(DBT)) %>% drop_na(Height) %>% nrow()` tree height-diameter observations in the dataset, with log-transformed diameter as the fixed effect, and tree nested within plot as random effect. This model explained `r round(MuMIn::r.squaredGLMM(htDiamFit)[1]*100,1)`% of variation in log-transformed heights, with an intercept of `r round(fixef(htDiamFit)[1],3)` and slope coefficient of  `r round(fixef(htDiamFit)[2],3)` * log(DBH).  Residual and predicted vs actual height plots were assessed to ensure goodness of fit (`r appNums("htModelFit",display="cite")`).  The model slightly overestimated tree heights in shorter trees, and underestimated tree heights in taller trees.  This model was applied to predict heights in trees without height measurements, and tree-volume estimates generated as per Nigh [-@nigh2016].

```{r hardcode volume coefficients per Nigh}

volCoef<-
  
  ## TOTAL VOLUME
  #  Subalpine fir 
  data.frame(Species="Bl",VolType="total",b0=-9.79,b1=1.813,b2=1.033) %>% 
  
  # Douglas-fir
  add_row(Species="Fd",VolType="total",b0=-10.19,b1=1.73,b2=1.206) %>% 
  
  # Add spruce
  add_row(Species="Sx",VolType="total",b0=-10.024,b1= 1.801,b2= 1.107) %>% 
  
  ## MERCHANTABLE VOLUME
  # Subalpine fir
  add_row(Species="Bl",VolType="merch",b0=-9.980,b1= 1.792,b2= 1.102) %>% 
  
  # Douglas-fir
  add_row(Species="Fd",VolType="merch",b0=-10.503,b1= 1.724,b2= 1.296) %>% 
  
  # Spruce
  add_row(Species="Sx",VolType="merch",b0=-10.280,b1= 1.769,b2= 1.206 ) %>% 
  
  mutate(BGC.Zone="SBS") %>% 
  
  # convert species to factor
  mutate(Species=factor(Species)) %>% 
  dplyr::select(Species,BGC.Zone,everything())

```



```{r generate volume estimates}

dat<-
  
  # Total volumes
  dat %>% 
  left_join(filter(volCoef,VolType=="total"),by="Species") %>% 
  mutate(VolTot=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>% 

  # Merchantable volume
  left_join(filter(volCoef,VolType=="merch"),by="Species") %>% 
  mutate(VolMerch=exp(1)^b0*DBH^b1*Height.Pred^b2) %>% 
  dplyr::select(-c(BGC.Zone,VolType,b0,b1,b2)) %>% 
  
  # basal area
  mutate(BA=(DBH/100/2)^2*pi) 

# Create stand data
standDat<-
  dat %>% 
  group_by(Plot,Year) %>% 
  summarise(Vol.Tot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
            Vol.Merch=sum(VolMerch,na.rm=T)*20,
            BA=sum(BA,na.rm=T)*20,
            k=n()*20) %>% 
  mutate(QMD=sqrt(BA/(k*0.0000785))) %>% 
  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting


```

```{r boxplot stand volumes by year}

volPlot<-
  standDat %>% 
  # plot
  ggplot()+
  aes(x=Year,y=Vol.Tot,fill=BA.Target)+
  geom_boxplot()+
  facet_wrap(~TU)+
  theme(legend.position="none")+
  ylab("Plot volume (m³/hectare)")

# caption
figNums("volPlot.plot",display=FALSE)

volFit<-
  standDat %>% 
  lm(Vol.Tot~BA.Target*Year,data=.)


```

  
### Stand increment
```{r function to compute stand increment}

standIncFunc<-function(varName) {


  standDat %>% 
  dplyr::select(Plot,Year,varName,BA.Target) %>% 
  pivot_wider(names_from="Year",values_from=varName) %>% 
  mutate(`1994Inc`=`1994`-`1992`) %>% # stand periodic increment 
  mutate(`1997Inc`=`1997`-`1994`) %>% # stand periodic increment 
  mutate(`2009Inc`=`2009`-`1997`) %>% # stand periodic increment 
  mutate(`2019Inc`=`2019`-`2009`) %>%  # stand periodic increment 
  
  # divide by # growing seassons in each period
  mutate(`1994Inc`=`1994Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`1997Inc`=`1997Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`2009Inc`=`2009Inc`/11) %>% # 11 growing seasons in this period 
  mutate(`2019Inc`=`2019Inc`/11) %>% 
    dplyr::select(Plot,BA.Target,`1994`=`1994Inc`,`1997`=`1997Inc`,
                  `2009`=`2009Inc`,`2019`=`2019Inc`) %>% 
    pivot_longer(cols=`1994`:`2019`,names_to="Year",values_to="Inc") %>% 
    mutate(varname=varName) %>% 
    dplyr::select(Plot,BA.Target,varname,Year,Inc) %>% 
    return() # return
}
  
```


```{r apply stand increment function}

incPlot<-
  standIncFunc("Vol.Tot") %>%   # extract total volume 
  rbind(standIncFunc("Vol.Merch")) %>%   # extract total volume 
  rbind(standIncFunc("BA")) %>% # extract basal area
  drop_na(Inc) %>% 
  
  # Create group summaries
  group_by(BA.Target,Year,varname) %>% 
  summarise(IncMean=mean(Inc),
            se=sd(Inc)/sqrt(length(Inc))) %>% 
  
  # create lookup column
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% # create lookup column
 
  # plotting
  ggplot()+
  aes(x=Year,y=IncMean,color=TU,group=1)+
  geom_errorbar(aes(ymin=IncMean-se,ymax=IncMean+se,width=.2))+
  geom_point(size=2)+
  geom_path()+
  facet_grid(varname~TU,scales="free_y")+
  theme(legend.position="none")+
  ylab("Annualized increment by period")

# caption
figNums("incPlot",display=FALSE)

```

Annual increment between the four periods differed between treatment units (`r figNums("incPlot",display="cite")`).  In the first three growing seasons after treatment, basal area and volume increments in the two treatment units with basal area removal were lower than increment in the control unit.  In the high-removal treatment unit, basal area increased in each subsequent measurement period, and exceeded the other units between 1997 and 2019.  In the low-removal and control treatment units, increment either declined or stabilized after ca.1997.  Volume increment declined in all units from 2009-2019, and in the control unit, annual volume increment was lower in 2019 than in 1992.  

### Species composition over time

```{r create stand data by species}
# Create stand data
sppDat<-
  dat %>% 
  group_by(Plot,Species,Year) %>% 
  summarise(Vol.Tot=sum(VolTot,na.rm=T)*20, # notice that all variables are by hectare
            Vol.Merch=sum(VolMerch,na.rm=T)*20,
            BA=sum(BA,na.rm=T)*20,
            SPH=n()*20) %>% 
  mutate(QMD=sqrt(BA/(SPH*0.0000785))) %>% 
  left_join(plotDat,by="Plot") %>%  # joinwith plot Data
  dplyr::select(Plot:BA.Target) %>% 
  pivot_longer(cols=Vol.Tot:QMD,names_to="varName",values_to="Quantity") %>% 
   mutate(TU=paste("TU",BA.Target,sep=":"))  # create a TU column for plotting

sppIncFunc<-function(varName) {


  sppDat %>% 
  dplyr::select(Plot,BA.Target,varName,Year) %>% 
  pivot_wider(names_from="Year",values_from=varName) %>% 
  mutate(`1994Inc`=`1994`-`1992`) %>% # stand periodic increment 
  mutate(`1997Inc`=`1997`-`1994`) %>% # stand periodic increment 
  mutate(`2009Inc`=`2009`-`1997`) %>% # stand periodic increment 
  mutate(`2019Inc`=`2019`-`2009`) %>%  # stand periodic increment 
  
  # divide by # growing seassons in each period
  mutate(`1994Inc`=`1994Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`1997Inc`=`1997Inc`/3) %>% # 3 growing seasons in this period 
  mutate(`2009Inc`=`2009Inc`/11) %>% # 11 growing seasons in this period 
  mutate(`2019Inc`=`2019Inc`/11) %>% 
    dplyr::select(Plot,Species,BA.Target,`1994`=`1994Inc`,`1997`=`1997Inc`,
                  `2009`=`2009Inc`,`2019`=`2019Inc`) %>% 
    pivot_longer(cols=`1994`:`2019`,names_to="Year",values_to="Inc") %>% 
    mutate(varname=varName) %>% 
    dplyr::select(Plot,Species,BA.Target,varname,Year,Inc) %>%
    mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
    return() # return
}


```


```{r plot species breakdown}
sppPlot<-
  sppDat %>% 
  filter(Species %in% c("Bl","Sx")) %>% 
  ggplot()+
  aes(x=Year,y=Quantity,fill=Species)+
  geom_boxplot()+
  facet_grid(varName~TU,scales="free_y")+
  theme(legend.position="bottom")

# caption
figNums("sppPlot",display=FALSE)

```

#### Volume and basal area
Fir basal area and volume exceeded that of spruce in both the control and light-removal treatment unit; these differences increased in each subsequent period after ca.1997 (`r figNums("sppPlot",display="cite")`).  From 1997-2019, fir basal area and volume increased faster than spruce in the light-removal and control units.  The differences were not as clear in the high-removal treatment unit, as there was a lot of variation in spruce among plots.  *we could test this for significance*

Among treatments, spruce volumes and basal were similar across the three treatment units, especially from 1997-2019.  Spruce volume and basal area varied substantially among plots within the high-removal treatment unit.  Fir basal area and volumes in the control unit exceeded other treatment units, with the largest difference between control and high-removal treatment unit.

#### Tree density
In the control and light-removal treatment unit, there were more fir trees than spruce across all time periods.  The difference between fir and spruce density appeared to remain stable in the control unit over the period of this study.  In the high-removal treatment unit, fir density began to noticeably exceed that of spruce starting ca.1997, although the variation in fir tree density among plots in this treatment was relatively high.  

#### Quadratic mean diameter
Quadratic mean diameter in both fir and spruce remained relatively stable over all measurement periods in the control unit.  In the low-removal treatment unit, spruce QMD exceeded that of fir across all periods, however, the differences between species apeared to be diminishing over time.  In the high-removal treatment unit, spruce QMD increased over time, and both species had approximately similar mean QMD.

```{r compare species proportions as per Deal 2002}

sppProp.Dat<-
  dat %>%
  filter(Year %in% c(1992,2019)) %>% # filter for start and end years
  group_by(Plot,Year) %>% 
  summarise(N.Plot=n()) %>% # summarise number of years per plot
  group_by(Plot) %>% 
  summarise(N.Plot=n()) %>% 
  filter(N.Plot==2) %>% 
  pull(Plot) %>% 
  data.frame(Plot=.) %>% 
  
  # now filter dat for these plots
  right_join(dat,by="Plot") %>% 
  filter(Year %in% c(1992,2019)) %>% # filter for years
  filter(Status=="Live") %>% #filter for live trees
  
  # break trees into diameter classes
  mutate(DBH.class=cut(DBH,breaks=c(7.5,12.5,18.5,30,90),include.lowest = TRUE,right=FALSE)) %>% 
  
  # Calculate plot BA, Volume and SPH by species, plot and DBH.class 
  group_by(Plot,Species,Year,DBH.class) %>%
   summarise(BA=sum(BA)*20,# Notice the plot multiplier
            Vol=sum(VolTot)*20,
            SPH=n()*20) %>% 
  
  # Calculate species proportions
    filter(Species %in% c("Bl","Sx")) %>% # filter for Sx and Bl
    pivot_longer(cols=BA:SPH,names_to="Var",values_to="Value") %>% 
 pivot_wider(names_from=Species,values_from=Value) %>% 
  replace_na(list(Bl=0,Sx=0))      # replace NA with 0


# Plotting
sppProp.Plot<-
  sppProp.Dat %>% 
  mutate(Total=Bl+Sx) %>% 
  mutate(Bl=Bl/Total,
         Sx=Sx/Total) %>% 
  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Prop") %>% 
  left_join(plotDat,by="Plot") %>% 
  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
  summarise(Prop=mean(Prop)) %>%
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
  filter(Var=="Vol") %>% 
  ggplot()+
  aes(x=DBH.class,y=Prop,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_grid(Year~TU)+
  ylab("Proportion")+
  xlab("Diameter class (cm)")+
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 270,vjust=0.2,hjust=0))


sppVol.Plot<-
  sppProp.Dat %>% 
  pivot_longer(cols=Bl:Sx,names_to="Species",values_to="Val") %>% 
  left_join(plotDat,by="Plot") %>% 
  group_by(BA.Target,Year,DBH.class,Var,Species) %>% 
  summarise(Val=mean(Val)) %>%
  mutate(TU=paste("TU",BA.Target,sep=":")) %>% 
  filter(Var=="Vol") %>% 
  ggplot()+
  aes(x=DBH.class,y=Val,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_grid(Year~TU)+
  ylab("Volume (m³/ha)")+
  xlab("Diameter class (cm)")+
  theme(legend.position="none",
        axis.title.x=element_blank(),
        axis.text.x=element_blank())
  

sppProp.plot<-plot_grid(sppVol.Plot,sppProp.Plot,nrow=2,rel_heights=c(1,1.5),axis="lr",align="h")

figNums("sppProp.plot",display=FALSE)
```

# Discussion
Partial cutting silviculture systems aim to affect the growth and structure of a stand primarily through the manipulation of stand density, tree spatial distribution, tree size distribution and species composition.  The growth and regeneration response of the residual stand will vary with type of tree removal (ie thinning from below or above, or mixed), residual stand structure, silvical characteristics of the residual species, site productivity, tree distribution, stand age and health, and other site-level factors.  We note that the literature contains ample studies of commercial thinning in even aged stands (references), however, studies of selection harvest systems in uneven-aged stands such as this study are relatively rare. 

Consistent with other studies, we found that tree-level growth increment (basal area and volume) and recruitment of new trees into various size classes varied by species, tree size at the time of treatment, and level of basal area removal.  These changes in growth and stand structure resulted in stand-level changes that suggest the following, specific to uneven-aged spruce-fir forests in our study region:

1. Higher levels of basal area removal can stimulate a stronger and more sustained growth response; 
2. Partial cutting can be used to maintain pre-harvest species composition (in other words, maintain proportions of spruce and fir that are similar to pre-harvest stand composition);
and 
3. Different levels of partial cutting can maintain a stand structure with trees well represented in different size classes (commonly referred to as uneven-aged stand structure).


## Growth responses to partial cutting
Partial cutting in mixed-species uneven-aged stands requires an understanding of species- and size-specific growth responses to different levels of stand manipulation [@Deal2002].  Our results suggest that a higher level of basal area removal (i.e., a residual basal area (RBA) of 10m^2^ in 1992) stimulated a stronger and more sustained tree-level growth (both volume and basal area increment) response than the high RBA treatment unit and the control unit.  

Over each period, basal area increment was higher in the low RBA treatment units compared to other two, and peaked ca. 20 years after treatment, compared to an increment peak of ca. 5 years in high RBA and control units.  At the tree-level, spruce trees grew faster than fir, and in both species, larger trees responded with more growth increment than smaller trees (although this was reversed when increment was expressed as a percentage of original tree size).  Longer periods (i.e., over 40 years) of increased growth rates have been documented after understorey release by insect attack in subalpine spruce-fir forests in Colorado, US [@Veblen1991].  


## Stand-level changes

A stand's cumulative volume over time can be estimated with stand density and periodic increment [@Pretzsch2005]. 

Spruce trees in the understorey can have a clumpier distribution than subalpine fir, which may indicate that spruce has a narrower range of suitable microsites for regeneration [@Shea1985; @Klinka2000].  On the other hand, subalpine fir can regenerate across a wider range of substrates, and therefore may be expected to proliferate more uniformly in the understorey [@Klinka2000].   Thus, partial harvesting systems that aim to maintain spruce composition in the stand should aim to protect spruce understorey as much as possible, as well as create suitable microsites for seed germination such as exposed mineral soil [@Coates1994].  Once established, the performance of planted spruce and subalpine may be more or less equal [@references].  Understorey Engelmann spruce and subalpine fir displayed similar growth responses to release after bark beetle attack in a subalpine forests in Colorado [@Veblen1991]. 


### Species- and tree-size factors influencing growth responses

Our results show that both species had initial delayed growth response after treatment.  Trees acclimated to shady conditions may suffer a shock after release due to relatively rapid changes in light, temperatures and humidity which can increase evapotranspiration and respiration stress [get sources from @Wang2011].  A post-release lag in height growth response may also reflect the tree's reallocation of of carbon to below-ground structures at the expense of aboveground, as per the functional balance theory [@Kneeshaw2002; @Brouwer1962], whereas a relatively immediate response in basal area increment response after treatment may represent a species strategy to thicken stem afer release in response to increased wind stress [@Wang2011].

Hybrid spruce and subalpine fir share similar silvics and life histories [@Klinka2000], and have demonstrated similar growth and survival in managed forests in British Columbia [@Lajzerowicz2006] and elsewhere [@Veblen1991].  Thus, we may have anticipated similar growth responses between these species.  

### Spruce

### Fir
In an overstory removal experiment near our study region, understorey subalpine fir height growth responded best over three years to partial canopy removal, whereas radial growth responded strongest under complete canopy removal [@Wang2011].   Subalpine fir height growth response after overstory thinning varied with tree size and level of removal; trees of all sizes responded with positive height growth responses after partial canopy removal, whereas small tree also responded with positive growth after complete removal, which conversely caused height growth reductions in both intermediate and large trees [@Wang2011].  Radial growth was highest under a complete overstory removal, especially in intermediate trees, after an initial lag [@Wang2011].    

### Tree size
Across both species and all treatments, post-harvest growth increment increased with residual tree size.  This likely reflects that larger trees simply put on more growth in terms of basal area or volume than smaller trees.  This also suggests that the trees in this study were not so large that they could not respond to changes in stand conditions [@reference].  
Intermediate crown-class trees may respond better to partial cutting, because the treatment will result in a greater shift in light conditions for these trees compared to larger trees occupying a dominant or co-dominant crown class[@Wang2011].  Intermediate trees also may have higher vigour than suppressed trees, with a greater corresponding ability to release after treatment [@Wang2011].


## Stand structure dynamics over time

All treatment units maintained a stand structure analagous to an uneven-aged stand (ie reverse-J diameter distribution).  In the low RBA unit, most recruitment was into the smallest size classes, whereas recruitment was into larger size classes in the high RBA and control units.  This suggests that understory response may increase with level of change in the canopy, and that  more substantial changes to the stand conditions can create more favourable conditions for understory suppressed and intermediate trees to release and increase size.  We did not account for trees smaller than 7.5cm DBH in the study, therefore, we do not have data on seedling germination and establishment after treatment.  THe increase of tree density in the lower size classes in the low RBA may have been a result of increased seedling establishment after harvest.  However, our data show that tree density increased in the smaller size classes in the low RBA unit.  Natural regeneration is complex and in some cases can be largely stochastic, which can make its use difficult to achieve management objectives [@Bataineh2013].  We note that in high-elevation spruce-fir stands in British Columbia, understorey tree age and diameter may be uncorrelated [@Puttonen1998].

Within a time period, spruce quadratic means diameters were inversely related to harvest level.  QMD was highest in the control unit, and lowest in the high-removal treatment unit, suggesting that there are fewer but larger trees in the control, whereas recuitment into sapling class (>7.5cm DBH) trees was higher in high-removal TU.  

We note that our data do not provide information on the effects of partial harvest removal on spatial distribution of tree stems, which may be an important consideration when applying silvicultural treatments [@Saunders2008].  Spatial characteristics include spatial distribution of tree stems (e.g., clumped or even distribution), the degree of species intermingling, and size differences among neighbours, all of which can influence resource availablility, regeneration success and competition [cf @Saunders2008].   

## Management implications
Partial harvesting 

Both treatments resulted in a stand that would be considered fully stocked immediately after harvest according to the BC free growing guidelines [@reference].

***
Draft material below this line

-	Our study was conducted in a relatively homogenous site (most plots are located on sites with circum-mesic soil edaphics) which represents the majority of managed forests in this region.  The application of partial harvesting on azonal sites should be guided by site-specific data, as site productivity was found to have an important (and potentially opposite) effect on stand dynamics after harvesting [@Pretzch2005].  
-	Stand-level periodic increment follows a unimodal density optimum pattern [@Pretzsch2005], which can be used by silviculturists to manipulate density in order to maximize periodic increment and cumulative volume.  Our results suggest that higher levels of basal area removal resulted in tree-level growth rate increases that were higher and more sustained than the high RBA treaetment unit and control.  After approximately 20 years, the low RBA unit had converged with the high RBA unit in terms of stand-level basal area and volume, suggesting a higher cumulative volume in the low RBA unit.  Presumably, an even higher level of removal (e.g., residual basal area of 5m2/ha) could have resulted in a reduced stand-level periodic increment and subsequent lower cumulative volume, as increased growth rates would be offset by lower stand density, reflecting the unimodal density optimum pattern [@Pretzsch2005].  This density optimum may also vary with stand age [@Pretzsch2005].  
-	By three decades after treatment, both treatment units still had lower volume than the control, however, the differences between the three units diminished over time.  Given enough time, stand volumes may converge.  A study in thinned Norway spruce stands found that cumulative volumes in thinned treatments exceeded unthinned volumes by 30-39 years after treatment [@Pretzsch2005]; however, it may not be suitable to compare these fiindings to our study, as the treatment was a thin from below.


Here are the main results that emerge to me from our data:

Silviculture treatments aim to affect the growth of stands through the manipulation of stand density, composition, distribution.  Effects of stand density and composition management on growth and yield will vary with type of treatment (from below or above), site productivity, silvics of the residual species, tree distribution (clumped, dispersed, etc.), stand age, forest health factors and other factors.

Partial removal of the overstorey can allow forest managers to manipulate light levels in the stand, however, the successful application of this treatment requires an understanding of how the understory trees will respond [@Wang2011].







Across species, removal intensity and size classes, trees showed a growth response after treatment.  


\newLine
\newLine


## Density-growth relationship
2. High-level of basal area removal stimulated a stronger and more sustained growth response
Basal area increment has higher in the high-removal treatment units compared to other two, and peaked ~20 years after treatment, compared to ~5 years in control/light-removal.  What does this mean for stand development over the next 20-30 years?  
- By 2019, the high-removal and light-removal TU converged in terms of volume and basal area.  This seems remarkable to me, and perhaps runs counter to what we learned in silviculture II: *one will never recoup thinned volume!*  Growth rates in the high-removal TU exceeded those in the light-removal; what does this mean for stand dynamics over the next 20 years?  Will the high-removal TU actually exceed light-removal in terms of volume?
- Most basal area increases are from the larger trees, however, the smaller trees had stronger responses when considered as a % of initial tree size.  

\newLine
\newLine

## Species differences in growth response


### Spruce




## Species composition
In a retrospective analysis of partially cut hemlock-sitka spruce stands in Alaska, partial cutting had little effect on species composition after treatment [@Deal2002].  
Partial cutting in this uneven-aged spruce-fir forest affected spruce-fir composition differently depending on harvest level, as well as tree sizes.
- Stand development after harvest showed that spruce volume, basal area, and QMD increased more in the high-removal treatment compared to the other two treatments.  In several plots in the high-removal TU, spruce volumes exceeds control/light-removal spruce. 
- High removal stimulated stronger growth responses in spruce trees, compared to fir. In the light-removal and control units, spruce basal area and volume changed much less over the 27 years of the trial, and in fact, have similar levels.  
- 
- Would we see this response in even more removal, say down to 5m^2^ per hectare?
- An important note: fir stems per hectare were low in the high-removal TU until around 1997, and then increased more rapidly relative to other units from ca. 1997-2019.
- Note that Figure 3 and 7 are similar - Figure 3 shows predicted BAI from model, and 7 shows actual data.  Recommend deleting Figure 3 and using 7, b/c prefer to use real data!  Appendix three is similar to Figure 7 - just uses boxplots instead of mean/se.  Recommend Figure 7 and delete Appendix 3.

\newLine
\newLine

4. What other results emerge from the data?

\newLine
\newLine

5. Management implications



**Discussion structure from Mike below**

1.  comparisons of findings to the related scientific literature
2.  comparisons of findings to biological principles and current ecological understandings - e.g. do our results make sense based on our understandings of stand dynamics, or example?
3.  discussion of findings that seem anomalous, counter-intuitive, or contrary to prevailing literature or theory, AND consideration of whether our results and data may be an advancement in the science.
4.  future research directions
5.  commentary and direction on our findings and current management practices relative to: (a) similar practices globally, (b) similar practices in similar Picea-Abies types elsewhere, and finally (c) practices in ESSF-SBS spruce fir types in BC.
6.  Regarding commentary on BC management:  Keep this squarely focused on what the scientific foundation of these types of practices SHOULD be, based on the scientific evidence (and comparative literature) we have assembled. We can then comment on current practices and standards and whether they do or do not meet these tests. We must acknowledge that management standards for silviculture incorporate multiple considerations including best available science, but also other factors (e.g. simplicity, enforceability, etc).
7.  Extrapolation and interpretation of our results beyond the bounds of this particular study and study site:  this is where the comparative literature and reference to established principles of stand dynamics and productivity will aid us - i.e. if our results are consistent with broad-based silvicultural and ecological understandings and principles, then we can have good confidence that our results are robust and sound for extrapolation of other similar stands and forest types.
8.   Opportunities for modelling etc in future?



# Tables

##### page break

# Figures

```{r plot marginal means DBH increment by tree size}
fitBA.plot
```

`r figNums("fitBA.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment by tree basal area at time of treatment, species and treatment unit.  Interactions between initial tree size, treatment unit and species were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBA)[1] %>% round(3)*100, "% variation in basal area increment (not shown).  Predictions are back-transformed to original response scale.",sep=""))`


```{r plot marginal means DBH increment pct}

fitBApct.plot

```

`r figNums("fitBApct.plot",caption=paste("Predicted basal area (m^2^ at breast height) increment (cm), as a percentage of tree basal area at the time of treatment, by initial tree size.  Interactions between initial tree size and (i) treatment unit and (ii) species factors were significant (p<0.001) in a linear model explaining ", MuMIn::r.squaredGLMM(fitBApct)[1] %>% round(3)*100, "% variation in basal area increment (not shown). Predictions are back-transformed to original response scale.",sep=""))`

##### page break

```{r plot marginal means BA increment by year}

fitBA.Ann.plot


```

`r figNums("fitBA.Ann.plot",caption=paste("Predicted annualized basal area increment (m^2^ at breast height) over four periods from a linear mixed effects model explaining ", MuMIn::r.squaredGLMM(fitBA.Ann)[1] %>% round(3)*100, "% variation in log-transformed basal area increment with period, treatment unit and species as fixed effects (p<0.001 for all fixed effects, now shown). Basal area increment are average for growing seasons within each period.  For example, values at x-axis value of 1994 are mean for three growing seasons (between 1992 and 1994).  Predictions are back-transformed to original response scale.  Error bars represent 95% confidence intervals.",sep=""))`

##### page break

```{r plot density by diameter class, fig.width=6}
sphPlot1

```

`r figNums(name="sphSpeciesPlot",caption="Mean density of stems per hectare (SPH) by diameter class and species, faceted by treatment unit and two measurement periods.  Lowest value is included in each diameter class, highest value is not.")`

##### page break

```{r plot density change PCT by diameter class, fig.width=6}
sphPCTPlot

```

`r figNums(name="sphPCTPlot",caption="Percentage change between 1992 and 2019 in stems per hectare (SPH) by diameter class, faceted by treatment unit and species.  Lowest value is included in each diameter class, highest value is not.")`

##### page break

```{r volume stand plots}
volPlot

```

`r figNums("volPlot.plot",caption="Variation in total volume per plot (m³ per hectare) at five measurement years, faceted by treatment unit.")`

##### page break

```{r plot stand increment}

incPlot

```

`r figNums("incPlot",caption="Mean annual increment per plot over four measurement periods.  Top panel row shows basal area (BA), with units of m² per hectare.  Middle panel row shows merchantable volume (Vol.Merch), and bottom panel row shows total volume (Vol.Tot).  Volume units are m³ per hectare.")`

##### page break
```{r plot stand attributes by species,fig.height=8,fig.width=6}
sppPlot

```

`r figNums("sppPlot",caption="Among-plot variation in stand attributes, by species and faceted by treatment unit and attribute.  Units are as follows: basal area (BA, m² per hectare); QMD (quadratic mean diameter,cm), trees per hectare (SPH, count), Merchantable volume (Vol.Merch, m³ per hectare); Total volume (Vol.Merch, m³ per hectare)")`

##### page break
```{r plot total volume species proportion,fig.height=6}

sppProp.plot


```

`r figNums("sppProp.plot",caption="Proportion of mean plot volume by four diameter classes of fir and spruce, expressed as volume (m³ per hectare, top panel) and proportion in bottom panel.  Almost identical proportions were obtained when comparing basal area and tree density.")`

# Appendices
```{r plot raw PBAI by species and TU}

pi.Init.rawPlot


```

`r appNums("piInitRaw",caption="Log-transformed basal area increment (m²) over 1992-2019 period by log-transformed tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.  Both axes have been log-transformed.  Linear regression lines added to emphasize slope.")`

##### page break

```{r plot raw PBAI as pct initial tree size by species and TU}

piPCT.Init.rawPlot


```

`r appNums("piPCTInitRaw",caption="Log-transformed basal area increment (m²), as a percentage of log-transformed initial tree size, over 1992-2019 period by tree size (basal area, m²) at the time of treatment, faceted by treatment unit and species.  Both axes have been log-transformed.  Linear regression lines added to emphasize slope.")`

##### page break

```{r plot raw basal area increment data by year}
baAnn.raw.plot

```


`r appNums("baAnnRaw",caption="Annualized basal area increment (m² at breast height) over four measurement periods. Boxplots show dispersion of mean growing season tree-level basal area increment within each period.  Data are faceted by species and treatment unit.")`

##### page break

```{r plot height to diameter models by species and year, fig.asp=1.2, fig.width=6}

heightDiamRaw.plot

```

`r appNums("heightDiamRawPlot",caption="Height and diameter comparisons for 870 tree measurements, faceted by species and measurement year.")`

##### page break
```{r plot diam to height model fit}

htModelFit.plot
```

`r appNums("htModelFit",caption="Predicted height against measured height for 870 tree measurements.  Outliers above the line may represent a tree with a broken top.")`


# References
