---
title: "Fdi basal area analysis"
author: "Hardy Griesbauer"
date: "06/03/2020"
output: html_document
---

## The mystery of the disappearing Fdi basal area
The figure below suggests that Douglas-fir basal area declined noticeably between 1992 and 1994 in the control and 20m2 residual treatment units.  Below is some script to analyze this phenomenon further.  

```{r setup,include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=8)

# Load libraries
library(tidyverse)

# Load dataset
load(here::here("data","ep1162_Data.RData"))
load(here::here("data","ep1162_plotData.RData"))

# set up diameter classes
diamClass=seq(from=7.5,to=95,by=5)


dat %>% 
    filter(Species!="Ep") %>% # remove birch 
  filter(Status=="Live") %>% # filter for live only trees
   left_join(plotDat,by="Plot") %>% # join with plot data
   mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) %>% # assign DBH classes
  mutate(BA=(DBH/100/2)^2*pi) %>% # convert tree DBH to basal area
  group_by(BA.Target,Plot,Year,Species) %>% 
  summarise(BA=sum(BA)*20) %>% # multiply by 20 for BA/ha
  ungroup() %>% 
  
  # create mean and SD
  group_by(BA.Target,Year,Species) %>% 
  summarise(BA.Mean=mean(BA),
            BA.se=sd(BA)/sqrt(length(BA))) %>% # calculate standard error of the mean
  
    # Add labeller column
  mutate(label=factor(paste("Trt:",BA.Target,sep=""))) %>% 
  
  # # pad out the year column
  # right_join(data.frame(Year=as.factor(1992:2019)),by="Year") %>% 
  
  # Plotting
  ggplot()+
  aes(x=Year,y=BA.Mean,group=Species,fill=Species)+
  geom_bar(stat="Identity")+
  facet_wrap(~label)+
  ylab("Mean basal area per plot (m2/ha)")+
  xlab("Measurement year")
  
```

First step, let's isolate plots with Fdi in the two treatment units:
```{r}
fdiDat<-
  dat %>% 
  filter(Species=="Fd") %>% # filter for Douglas-fir entries
  left_join(plotDat,by="Plot") %>% 
  mutate(DBH.class=cut(DBH,diamClass,ordered_result = TRUE,right=F)) # assign DBH classes

```

Looks like there are two trees that are missing measurements from 1994 and 1997.

```{r}
fdiDat %>% 
  filter(TreeID %in% c(400,188))


```

Are there any other trees (non-Fd) that are missing mid-point measurements?

```{r}

  
  dat %>% 
  filter(TreeID!="New") %>% # have to remove these from dataset
  dplyr::select(Plot,TreeID,Year,DBH) %>% # select columns to run analysis on
  pivot_wider(names_from=Year,values_from=DBH) %>% 
  filter(!is.na(`1992`) & !is.na(`2019`)) %>% # filter for trees with measurements at 1992 and 2019
  pivot_longer(cols=3:7,names_to="Year",values_to="DBH") %>% 
  drop_na(DBH) %>% 
  group_by(Plot,TreeID) %>% 
  summarise(num=n()) %>% 
  filter(num<5) 
  
  
  

```


I suggest we interpolate these using linear regression:

```{r}





